Social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{Paper presented at the session 'Health and Mortality in Arctic Populations'. The Ninth International Congress on Arctic Social Sciences, Umeå, June 8--12, 2017}

### 2017-05-22 icass.Rnw
Update region description

\date{May 18, 2017}

@@

For a detailed presentation we refer to \citet{segb16}. Here is a short
summary.

@@

Both regions are part
of the county of Västerbotten in the north of Sweden, along the coast of
the Gulf of Bothnia. The Skellefteå region in Poplink consists of a selection
of parishes surrounding the town of Skellefteå, founded in 1845 but with a
very small population in the 19th century. The data from the period after
1960 cover the Skellefteå and Malå municipalities, the same area as for the
earlier period but with the addition of two more parishes. The majority
of the 19th century population lived in rural villages and hamlets, getting
their livelihood from agricultural production. The region was vulnerable for
harvest failures, one living long in memory was the famine of 1867–-1868
(Edvinsson and Broström, 2014). Towards the end of the century, the region became connected to the Swedish railway system, thus improving the communications and contacts with other parts of the country. During the
20th century, industrialisation took place, partly by the opening of mines in
the inlands. This also led to a population increase both in the town and in
the rural parts, and a much more diversified economy.
Mortality was fairly low in comparison with other parts of the country
and the fertility transition was late. The Skellefteå population size as defined
in our data sets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900,
61\,938 on January 1, 1950 and 76\,723 at the end of the century.

The Umeå region in Poplink (from 1900 until 1950s) consists of Umeå urban and
rural parish, and in the Linneus database from 1960 onwards of Umeå municipality where another three parishes are included. This region has a somewhat different character from that of Skellefteå, even though agriculture dominated the economy for a long time. Umeå town had a
small population but was substantially larger than Skellefteå town during almost
the whole studied period. It was the administrative centre in the county of
Västerbotten and schools and military regiments were placed here. Consequently, Officials,
teachers and the military were much more common in Umeå than in Skellefteå. Agriculture
dominated the rural part, there were some foundries, some
industries for example in forestry and small-scale production. Umeå had a
much more diversified occupational structure than Skellefteå. The population size
as defined in our data sets was 18\,970 on January 1, 1900, 32\,900 on
January 1, 1950 and 103\,970 when the century ended.

>>>>

\date{May 22, 2017}

@@

@@

The Skellefteå and Umeå regions are part of county of Västerbotten in the north of Sweden. This was a remote part of the country, where communications with the rest of Sweden were restricted until the late 20th century. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen. Communications during winter used the winter roads. The first ships during the year came to Umeå in May? Towards the end of the 19th century, railways were built up to our regions, thereby facilitating contacts with the rest of Sweden. But of course, they were not totally outside what happened in other parts of the country. The inlands were however very remote. But our regions are situated mostly close to the coast. During the 19th century, these parts became integrated in the same epidemiological pattern as in the rest of Sweden. Diseases reached to the places that earlier were saved.
Both regions are part of the county of Västerbotten in the north of Sweden, along the coast of the Gulf of Bothnia. The Skellefteå region in Poplink consists of a selection of parishes surrounding the town of Skellefteå, founded in 1845 but with a very small population in the 19th century. The data from the period after 1960 cover the Skellefteå and Malå municipalities, the same area as for the earlier period but with the addition of two more parishes. The majority of the 19th century population lived in rural villages and hamlets, getting their livelihood from agricultural production. The region was vulnerable to harvest failures, one living long in memory was the famine of 1867–-1868 (Edvinsson and Broström, 2014). Towards the end of the century, the region became connected to the Swedish railway system, thus improving the communications and contacts with other parts of the country. During the 20th century, industrialisation took place, partly by the opening of mines in the inlands. This also led to a population increase both in the town and in the rural parts, and a much more diversified economy.

Mortality was fairly low in comparison with other parts of the country and the fertility transition was late. The Skellefteå population size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900, 61\,938 on January 1, 1950 and 76\,723 at the end of the century.

The Umeå region in Poplink (from 1900 until 1950s) consists of Umeå urban and rural parish, and in the Linneus database from 1960 onwards of Umeå municipality where another three parishes are included. This region has a somewhat different character from that of Skellefteå, even though agriculture dominated the economy for a long time. Umeå town had a small population but was substantially larger than Skellefteå town during almost the whole studied period. It was the administrative centre in the county of Västerbotten and schools and military regiments were placed here. Consequently, officials, teachers and the military were much more common in Umeå than in Skellefteå. Agriculture dominated the rural part, there were some foundries, some industries for example in forestry and small-scale production. Umeå had a much more diversified occupational structure than Skellefteå. The population size as defined in our data sets was 18\,970 on January 1, 1900, 32\,900 on January 1, 1950 and 103\,970 when the century ended.
